|

### 2021-12-29 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_A.Rmd
revision to appendix

This appendix shows a basic workflow to fit a series of increasingly complex GAMs to simulated data that follows the trends presented in Section 5.1 in the main manuscript. Graphical and parameter diagnostics for goodness of fit are discussed, as well as model comparison via AIC (Aikake Information Criterion). For simplicity, the confidence intervals (CIs) shown in this section for the models are the across the function CIs created by `mgcv` by default. However, for the pairwise comparisons of the third model we use _simultaneous intervals_ as in the main manuscript, and the code for creating simultaneous CIs for the smooths can be found in **Appendix XXX**.

@@

@@

#alpha for ribbon in the smooth plots
al<-0.8

thm1<-scale_fill_scico_d(palette="tokyo",begin=0.3, end=0.8, direction = -1, aesthetics = c("colour","fill"))

@@

```{r,data, echo= TRUE}
dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
            Day=rep(c(0,2,5,7,10),times=2),
            Group=as.factor(rep(c("Control","Treatment"),each=5))
            )
```

@@

```{r,simulation}
n <- 10 #number of observations
sd <- 10 #approximate sd from paper
source(here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","simulate_data.R"))
dat_sim <- simulate_data(dat, n, sd)
```

@@

The first model fitted to the data is one that only accounts for different smooths by day. The model syntax specifies that `gam_00` is the object that will contain all the model information, and that the model attempts to explain changes in `StO2_sim` (simulated $\mbox{StO}_2$) using a smooth per `Day`. The model will use 5 basis functions (`k=5`) for the smooth. The smooth is constructed by default using thin plate regression splines. The smoothing parameter estimation method used is the restricted maximum likelihood (`REML`).

@@

gam_00<-gam(StO2_sim ~ s(Day, k = 5),
            method='REML',
            data  = dat_sim)

@@

```{r, first-GAM-diag, fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:first-GAM-caption)'}

appr1<-appraise(gam_00)
sm1<-draw(gam_00)
visual_check<-appr1+sm1

visual_check+plot_layout(design=layout1)
```

@@

```{r,gam-00-diagnostics, echo=TRUE, include=TRUE}
source(here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","gam_diagnostics.R"))
gam_diagnostics(gam_00)

@@

And now obtain a summary of the fitted model, which is obtained by calling `summary`.

@@

From the output of `gam_diagnostics`, it can be seen that the `k-index` is 0.35, which indicates that the model is not capturing the variability in the data. The `edf` (effective degrees of freedom) is an indicator of the complexity of the smooth. Here the complexity of the smooth is comparable to that of a 4th degree polynomial.

From the `summary` function, information about the assumed distribution of the errors (Gaussian in this case) and the link function can be obtained. The link function is 'identity' as the model does not make any transformation on the predictors. The 'significance of smooth terms' _p-value_ indicates if each smooth is adding significance to the model. Here, the _p-value_ is low but we have seen that there are issues with the model from the previous outputs. Finally, the 'deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to $\approx$ 12%.

@@

gam_01<-gam(StO2_sim ~ s(Day, by=Group,k = 5),
            method ='REML',
            data  = dat_sim)

source(here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","gam_diagnostics.R"))
gam_diagnostics(gam_01)
```

@@

```{r, second-GAM-diag, fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:second-GAM-caption)'}
#need to add figure number and caption
appr2<-appraise(gam_01)
sm2<-draw(gam_01)

visual_check2<-appr2+sm2

visual_check2+plot_layout(design=layout1)
```

@@

gam_02 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
            method='REML',
            data  = dat_sim)

source(here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","gam_diagnostics.R"))

@@

By using `appraise()` and `draw` on this model (Figure  \@ref(fig:final-GAM-diag)) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group. From `gam_diagnostics`, the k-index is now at an acceptable value ($\approx$ **1.23**).

@@

appr3<-appraise(gam_02)
sm3<-draw(gam_02)

visual_check3<-appr3+sm3

visual_check3+plot_layout(design=layout2)
```

@@

AIC(gam_00,gam_01,gam_02)

@@

The output in this case is expected: model `gam_02` has a lower AIC **(710.54)** whereas the initial two models have higher AICs (**884.43** and **805.30**). The AIC should not be considered as the only estimator of model quality, instead to be used as complimentary information to the graphical diagnostics and model checks described above.

@@

We want to emphasize that for through the manuscript and for the comparisons on `gam_02`, we included the group means in order to keep the pairwise comparisons on the scale of the response. By default, pairwise comparisons available in other software packages (such as in G. Simpson's _gratia_) do not include the group means. However, we decided to include them because for our approach, it is easier to see the magnitude in the change of the difference between two treatment groups when the means are included.

@@

source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","pointwise_comparisons.R"))
source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","difference_smooths.R"))
source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","pairwise_limits.R"))
#compute difference between smooths and calculate confidence interval: complete data
diff_complete <- difference_smooths(gam_02, smooth = "s(Day)", newdata = newdat,
                     unconditional = TRUE, frequentist = FALSE,
                     n=100, partial_match = TRUE, nrep=10000)
my_list<-pairwise_limits(diff_complete)
```

@@

```{r,pairwise-comp-workflow-B,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
rib_col<-'#8D7D82' #color for ribbon for confidence interval
control_rib <- '#875F79' #color for ribbon for control region
treat_rib <- '#A7D89E' #color for ribbon treatment region

c1 <- ggplot() +
  geom_line(data = diff_complete, aes(x = Day, y = diff),size=1, alpha=0.5) +
  annotate("rect",
              xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
              fill=control_rib,
              alpha = 0.5,
              )+
  annotate("text",
           x=1.5,
           y=-18,
           label="Control>Treatment",
         size=6,
         angle=90
         )+
    annotate("rect",
             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
             fill=treat_rib,
             alpha = 0.5
             ) +
  annotate("text",
             x=6,
             y=-18,
             label="Treatment>Control",
             size=6,
           angle=90
           )+
  geom_ribbon(data = diff_complete, aes(x = Day, ymin = lower_s, ymax = upper_s),
              alpha = 0.5, fill = rib_col, inherit.aes = FALSE) +
  geom_hline(yintercept = 0, lty = 2, color = "red")+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
    labs(y="Difference\n(Complete observations)")+
    theme_classic()+
    theme(
    axis.text=element_text(size=22))

@@

We also want to indicate that one of the major advantages of fitting GAMs in R using _mgcv_ is that once the appropriate conditional distribution has been chosen and model diagnostics are assessed, the user only needs to choose a basis dimension (the number of basis functions to use), and check if the basis dimension is adequately capturing the trend of the data using `k.check`. If that is  not the case, then the user needs to increase `k` a bit and check `k.check` again. The process might seem mechanic, but we hope that with the theory presented in the main manuscript and the workflow of this Appendix the user has a good understanding of what the model is doing and the rationale for choosing a GAM as a tool for statistical analysis.

>>>>

This appendix shows a basic workflow to fit a series of increasingly complex GAMs to simulated data that follows the trends presented in Section 5.1 in the main manuscript. Graphical and parameter diagnostics for goodness of fit are discussed, as well as model comparison via AIC (Aikake Information Criterion). For simplicity, the confidence intervals (CIs) shown in this section for the models are the across the function CIs created by `mgcv` by default. However, for the pairwise comparisons of the third model we use _simultaneous intervals_ as in the main manuscript, and the code for creating simultaneous CIs for the smooths can be found in Appendix B.

@@

```{r, include = FALSE}
knitr::opts_chunk$set(tidy = TRUE)
```

@@

#alpha for ribbon in the smooth plots
al <- 0.8

thm1 <- scale_fill_scico_d(palette = "tokyo", begin = 0.3, end = 0.8,
                         direction = -1, aesthetics = c("colour", "fill"))

@@

```{r,data, echo= TRUE}
dat<-tibble(StO2  = c(4, 27, 3, 2, 0.5, 7,4, 50, 45, 56),
            Day   = rep(c(0, 2, 5, 7, 10), times = 2),
            Group = as.factor(rep(c("Control", "Treatment"), each = 5)))
```

@@

```{r,simulation}
n <- 10 #number of observations
sd <- 10 #approximate sd from paper
source(here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final",
            "scripts", "simulate_data.R"))
dat_sim <- simulate_data(dat, n, sd)
```

@@

The first model fitted to the data is one that only assumes a single smooth over time shared by both the treatment and control groups The model syntax specifies that `gam_00` is the object that will contain all the model information, and that the model attempts to explain changes in `StO2_sim` (simulated $\mbox{StO}_2$) using a smooth per `Day`. The model will use four basis functions (`k = 5`) for the smooth. The smooth is constructed by default using thin plate regression splines. The smoothing parameter estimation method used is the restricted maximum likelihood (`REML`).

@@

gam_00<-gam(StO2_sim ~ s(Day, k = 5),
            method   ='REML',
            data     = dat_sim)

@@

The following code chunk generates graphical checks for the fitted GAM model with a single shared smooth shared between the treatment and control groups.

```{r, first-GAM-diag, fig.width=10, fig.height=7, out.width='75%', fig.align='center', message=FALSE, warning=FALSE,fig.cap='(ref:first-GAM-caption)'}

appr1 <- appraise(gam_00)
sm1 <- draw(gam_00)
visual_check <- appr1 + sm1

visual_check + plot_layout(design = layout1)
```

@@

```{r,gam-00-diagnostics, echo=TRUE, include=TRUE}
source(here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final",
            "scripts", "gam_diagnostics.R"))
gam_diagnostics(gam_00)

@@

From the output of `gam_diagnostics`, it can be seen that the `k-index` is 0.35, which indicates that the model is not capturing the variability in the data. The `edf` (effective degrees of freedom) is an indicator of the complexity of the smooth. Here the complexity of the smooth is comparable to that of a 4th degree polynomial. And now obtain a summary of the fitted model, which is obtained by calling `summary`.

@@

From the `summary` function, information about the assumed distribution of the errors (Gaussian in this case) and the link function can be obtained. The link function is 'identity' as the model does not make any transformation on the predictors. The 'significance of smooth terms' _p-value_ indicates if each smooth is significantly different from a constant mean 0 under the model. Here, the _p-value_ is low but we have seen that there are issues with the model from the previous outputs. Finally, the 'deviance explained' indicates how much of the variation in the data the model is able to capture, which in this case corresponds to $\approx$ 12%.

@@

gam_01<-gam(StO2_sim ~ s(Day, by=Group, k = 5),
            method ='REML',
            data   = dat_sim)

gam_diagnostics(gam_01)
```

@@

```{r, second-GAM-diag, fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:second-GAM-caption)'}
#need to add figure number and caption
appr2 <- appraise(gam_01)
sm2 <- draw(gam_01)

visual_check2 <- appr2 + sm2

visual_check2 + plot_layout(design = layout1)
```

@@

gam_02 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
            method ='REML',
            data   = dat_sim)

@@

By using `appraise()` and `draw` on this model (Figure  \@ref(fig:final-GAM-diag) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably close to an normal distribution, and the smooths are capturing the trend of the data within each group. From `gam_diagnostics`, the k-index is now at an acceptable value ($\approx$ 1.23).

@@

appr3 <- appraise(gam_02)
sm3 <- draw(gam_02)

visual_check3 <- appr3 + sm3

visual_check3 + plot_layout(design = layout2)
```

@@

AIC(gam_00, gam_01, gam_02)

@@

The output in this case is expected: model `gam_02` has a lower AIC (710.54) whereas the initial two models have higher AICs (884.43 and 805.30). The AIC should not be considered as the only estimator of model quality, instead to be used as complimentary information to the graphical diagnostics and model checks described above.

@@

We want to emphasize that throught the manuscript and for the comparisons on `gam_02`, we included the group means in order to keep the pairwise comparisons on the scale of the response. By default, pairwise comparisons available in other software packages (such as in G. Simpson's _gratia_) do not include the group means. However, we decided to include them because for our approach, it is easier to see the magnitude in the change of the difference between two treatment groups when the means are included.

@@

source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", 
                  "scripts","pointwise_comparisons.R"))
source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final",
                  "scripts", "difference_smooths.R"))
source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", 
                  "scripts", "pairwise_limits.R"))
#compute difference between smooths and calculate confidence interval: complete data
diff_complete <- difference_smooths(gam_02, smooth = "s(Day)", 
                                    newdata = newdat, 
                                    unconditional = TRUE,
                                    frequentist = FALSE,
                                    n = 100, partial_match = TRUE, 
                                    nrep=10000)
my_list <- pairwise_limits(diff_complete)
```

@@

```{r,pairwise-comp-workflow-B,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
rib_col <- '#8D7D82'     #color for ribbon for confidence interval
control_rib <- '#875F79' #color for ribbon for control region
treat_rib <- '#A7D89E'   #color for ribbon treatment region

c1 <- ggplot() +
  geom_line(data = diff_complete, aes(x = Day, y = diff), 
            size = 1, alpha = 0.5) +
  annotate("rect", xmin = my_list$init1, xmax = my_list$final1,
           ymin = -Inf, ymax = Inf, fill = control_rib,
           alpha = 0.5) + 
  annotate("text", x = 1.5, y = -18, label = "Control>Treatment",
           size = 6, angle = 90) +
  annotate("rect", xmin = my_list$init2, xmax = my_list$final2,
           ymin = -Inf, ymax = Inf, fill = treat_rib,
           alpha = 0.5) +
  annotate("text", x = 6, y = -18, label = "Treatment>Control",
           size = 6, angle = 90) +
  geom_ribbon(data = diff_complete, 
              aes(x = Day, ymin = lower_s, ymax = upper_s),
              alpha = 0.5, fill = rib_col, inherit.aes = FALSE) +
  geom_hline(yintercept = 0, lty = 2, color = "red") +
  scale_x_continuous(breaks = c(0, 2, 5, 7, 10)) +
  labs(y = "Difference\n(Complete observations)") +
  theme_classic()+
  theme(axis.text = element_text(size = 22))

@@

We also want to indicate that one of the major advantages of fitting GAMs in R using _mgcv_ is that once the appropriate conditional distribution has been chosen and model diagnostics are assessed, the user only needs to choose a basis dimension (the number of basis functions to use), and check if the basis dimension is adequately capturing the trend of the data using `k.check`. If that is  not the case, then the user needs to increase `k` a bit and check `k.check` again. The process might seem mechanistic, but we hope that with the theory presented in the main manuscript and the workflow of this Appendix the user has a good understanding of what the model is doing and the rationale for choosing a GAM as a tool for statistical analysis.
